\documentclass{beamer}
\usepackage{epsfig}
\title{An introduction to the {\bf Aa} language}
\author{Madhav P. Desai\\ Department of Electrical Engg.\\ IIT Bombay\\ Mumbai, India}

\begin{document}
\maketitle



\frame[containsverbatim]{\frametitle{The {\bf Aa} language}

\begin{itemize}
\item Serves as an intermediate representation in the AHIR-V2 flow.
\item Control-flow (imperative) language with support for parallelism
and branching.
\end{itemize}
}


\frame[containsverbatim]{\frametitle{A simple program in {\bf Aa}}
\begin{verbatim}
$module [maxOfTwo] $in (a: $uint<32> b:$uint<32>)
   $out (c: $uint<32>) $is
{
   c := ( $mux (a > b) a b )
}
\end{verbatim}
}


\frame[containsverbatim]{\frametitle{Program-structure in {\bf Aa} }
\begin{verbatim}
aA_Program :=  ( aA_Module | 
                 aA_Object_Declaration | 
                 aA_Named_Type_Declaration)*
\end{verbatim}
}

\frame[containsverbatim]{\frametitle{A module}
\begin{verbatim}
aA_Module := $module [identifier]
                $in ( aA_Module_Args) 
                $out (aA_Module_Args) 
{
        aA_Object_Declaration*
        aA_Atomic_Statement_Sequence
}
\end{verbatim}
}


\frame[containsverbatim]{\frametitle{{\bf Aa} Types}

{\bf Aa} provides a comprehensive set of types.
\begin{itemize}
\item {\em Unsigned integers} 
\begin{verbatim}
$uint<1>, $uint<32> etc.
\end{verbatim}
\item {\em Signed integers}   
\begin{verbatim}
$int<1>, $int<32> etc.
\end{verbatim}
\item {\em Sized floats}      
\begin{verbatim} 
$float<8,32>, $float<11,52>
\end{verbatim}
\item {\em Pointers} : 
\begin{verbatim}
$pointer<$uint<32>> etc..
\end{verbatim}
\item {\em Arrays}:  
\begin{verbatim}
$array[32][4] $of $uint<32> etc.
\end{verbatim}
\item {\em Records}: 
\begin{verbatim}
$record $uint<32> $uint<1>
\end{verbatim}
\item {\em Named Records}: 
\begin{verbatim}
$record [MyRec] $pointer<MyRec>) $uint<32>
\end{verbatim}
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Objects}
\begin{itemize}
\item Storage objects:
\begin{verbatim}
$storage A: $array [1024] $of $uint<32>
\end{verbatim}
\item Constant objects:
\begin{verbatim}
$constant A: $uint<4> := _b0011
$constant B: $uint<32> := _hf0f0f0f0
$constant C: $float<8,23> := _f2.3465e+0
\end{verbatim}
\item Pipe objects:
\begin{verbatim}
$pipe A: $uint<32> $depth 4
$lifo $pipe B: $uint<8> $depth 8
$noblock $pipe C: $uint<39> $depth 2
\end{verbatim}
\item Implicit objects:  these are defined by
statements:
\begin{verbatim}
a := (A + B)
\end{verbatim}
They are also called {\em static-single-assignment}
or SSA variables.
\item Interface objects:  Inputs and outputs of modules are
treated like SSA variables.  Inputs cannot be written into.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Storage Objects}
\begin{itemize}
\item  Will be implemented in memory spaces.
\item  Access to a storage object takes 3+ cycles.
\item  Should use sparingly.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Pipe Objects}
\begin{itemize}
\item Blocking in nature:  provide synchronization mechanism.
\item FIFO or LIFO data transfer within and across modules.
\item Non-blocking pipes: return a 0 if read is attempted on an empty pipe.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} SSA Objects}

\begin{itemize}
\item Implemented as registers.
\item Fast access (immediate read, single cycle write).
\item Should use as often as possible.
\item Type of these objects is inferred from the
context in which they appear.
\begin{verbatim}
a := (A + B)
\end{verbatim}
$A$ and $B$ must have the same type, and the type of $a$ is
inferred to be the type of $A$.
\end{itemize}

}


\frame[containsverbatim]{\frametitle{{\bf Aa} Expressions}
\begin{itemize}
\item Constants: 
\begin{verbatim}
_b00011
_habf1
\end{verbatim}
\item Simple references:
\begin{verbatim}
a
\end{verbatim}
\item Array references:
\begin{verbatim}
a[0][1] 
a[(I+1)][J][K] 
\end{verbatim}
\item Unary expressions:
\begin{verbatim}
(<op> expr)
e.g.  (~ a)
\end{verbatim}
\item Binary expressions:
\begin{verbatim}
(a <op> B)
<op> can be +,-,*,/,<<,>>,|,&,~|,~&,^,~^
            ==,!=,>,>=,<,<=
e.g. (a + b)
\end{verbatim}
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} More Expressions}
\begin{itemize}
\item Ternary expressions:
\begin{verbatim}
($mux <test-expr> <if-expr> <else-expr> )
e.g.  ($mux (a > 0) (b+1) (b-1))
\end{verbatim}
\item Concatenation expression:
\begin{verbatim}
(a && b)
\end{verbatim}
\item Bit-select expression:
\begin{verbatim}
(a [] I)
\end{verbatim}
\item Address-of expression:
\begin{verbatim}
@(a)
@(a[I])
\end{verbatim}
\item Pointer-dereference expression:
\begin{verbatim}
->(ptr)
\end{verbatim}
If it appears on the left-hand-side, it is
a store, else it is a load.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Complex Expressions}
\begin{verbatim}
(a [] I)
($slice a 9 4)
($concat a b c d)
($reduce | a b c d)
\end{verbatim}

}

\frame[containsverbatim]{\frametitle{{\bf Aa} Statements}
\begin{itemize}
\item {\em Atomic} statements.
\item {\em non-Atomic} statements.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{Atomic {\bf Aa} Statements}
\begin{itemize}
\item {\em Simple} statements.
\item {\em Block} statements.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Atomic Simple Statements}
\begin{itemize}
\item Assignment statements:
\begin{verbatim}
target-expression := source_expression
e.g.
a := (b + c)
\end{verbatim}
\item Call statements:
\begin{verbatim}
$call fpadd32 (A B) (C)
\end{verbatim}
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Atomic Simple Statements}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Atomic Block Statements}
\begin{itemize}
\item Series-block statements.
\item Parallel-block statements.
\item Branch-block statements.
\item Fork-block statements.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Series Block Statements}
\begin{verbatim}
$seriesblock [SB] {
   $storage a: $uint<32>
   a := (b + c)
   d := (a + e)
} (d => D)
\end{verbatim}

Control-flow is sequential: statements are executed in order, token
leaves statement after last statement finishes.  
\vspace{0.5cm}
A module body is also a series-block.
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Parallel Block Statements}
\begin{verbatim}
$parallelblock [PB] {
   a := (b + c)
   p := (q + r)
}
\end{verbatim}
Control-flow: both statements started in parallel,  token leaves
statement after both statements have finished.
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Branch Block Statements}
\begin{verbatim}
$branchblock [BB] {
   $merge $entry loopback
       $phi I := ($bitcast($uint<32>) 0) $on $entry
                    NI $on loopback
   $endmerge
   a[I] := (b[I] + c[I])
   NI := (I+1)
   $if (NI < 16) $then 
      $place [loopback] 
   $endif
}
\end{verbatim}
Control-flow:  sequential, but control flow is altered by
merge, place, if and switch statements.
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Phi Statements}
\begin{itemize}
\item Can appear only in branch blocks, and within merge statements.
\item Indicate the action to be taken depending on how control got
to this statement.
\item  Example:
\begin{verbatim}
       $phi I := ($bitcast($uint<32>) 0) $on $entry
                    NI $on loopback
\end{verbatim}

\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Do-Pipeline-While Statements}
These are not atomic, and can occur only inside a branch-block.
\begin{verbatim}
$dopipeline $depth 7 $buffering 1 $fullrate
   $merge $entry $loopback
        $phi I := ($bitcast ($uint<32>) 0) 
                      $on $entry
                   NI $on $loopback
   $endmerge
   a[I] := (b[I] + c[I])
   NI   := (I+1)
$while (I < 16)
\end{verbatim}
Control-flow:  sequential, controlled by the condition check.  The
places \$entry and \$loopback are implicitly defined.  When control
enters the do-while, the token gets placed in \$entry and when
control loops-back from the condition check, the token gets placed
in \$loopback.  The compiler will pipeline the loop by keeping
\$depth iterations alive.
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Fork Block Statements}
\begin{verbatim}
$forkblock [FB] {
   $seriesblock [S1] { ... }
   $seriesblock [S2] { ... }
  
   $join S1 S2 $fork
      $seriesblock [S3] { ... }
      $seriesblock [S4] { ... }
   $endjoin
 
   $join S3 S4 $endjoin
}
\end{verbatim}
Control-flow:  all statements will start in parallel, join-forks
will trigger new statements etc.  Token exits block when all statements
finish.
}



\frame[containsverbatim]{\frametitle{{\bf Aa} Volatile Statements}
\begin{itemize}
\item If you do not want to use a register:
\begin{verbatim}
$volatile p := (q + r)
\end{verbatim}
\item $p$ will not necessarily be implemented as a register.   You
can only assume that $p$ is equivalent to $(q  +  r)$.
\item Useful in describing combinational logic.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Summary}

\begin{itemize}
\item An {\bf Aa} program consists of a collection of module
descriptions and object declarations.
\item Each module is a sequence of statements.
\item Statements can be simple or block-structured.
\item Specific blocks allow branching, while others describe
parallelism.
\item Objects in an {\bf Aa} program can be FIFOs (pipes), storage
or implicitly inferred.
\item Almost everything you can express in C can be expressed, but
we are closer to hardware.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{Look at the FIR again}
}


\end{document}
